Is essentiality of genes conserved in Enterobacteriaceae?

### 2016-10-23 documents/manuscript.tex
Developed a new protein clustering algorithm using MCL

\subsection{Are there biases in transposon mutagenesis data?}
To evaluate the essentiality of a gene, the number of insertions within that gene was measured as explained in~\ref{sec:essentiality}. However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis. Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}. We performed a detailed study of these biases. The biases that we have studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.

@@

\subsubsection{Preferred insertion motif bias}
Another concern while inferring essentiality from transposon mutagenesis data is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, instead of their essentiality level. For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome. The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.

@@

We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}. The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes. However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.

@@

We needed to cluster sets of orthologous genes in the bacteria that we were studying. Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous. Plenty of methods are proposed for this purpose. Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall. We used Hieranoid for clustering orthologous genes.

@@

@@

For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database. This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2. Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class. Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.

>>>>

\subsection{Are there biases in transposon mutagenesis data?}
To evaluate the essentiality of a gene, the number of insertions within that gene was measured as explained in~\ref{sec:essentiality}. However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis. Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, kimura_nucleoid_2016}. We performed a detailed study of these biases. The biases that we have studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.

@@

\subsubsection{Preferred insertion motif bias}
Another concern while inferring essentiality from transposon mutagenesis data is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, instead of their essentiality level. For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome. The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases. \textcolor{red}{\{Our genomes are G-C rich, so it makes sense to see more G-Cs in this plot. Use BLogo\}}

@@

We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}. The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes. However, as mobile genetic elements are not stored in KEGG database \cite{kanehisa_kegg:_2000}, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.

@@

We needed to cluster sets of orthologous genes in the bacteria that we were studying. Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous. Plenty of methods are proposed for this purpose. Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall. We used Hieranoid for clustering orthologous genes. Hieranoid need a species tree for clustering genes. We used the method explained in Section~\ref{sec:phylogenetic} to generate this tree.

@@

\subsection{Generating the species tree}\label{sec:phylogenetic}
We first ran search and align commands from PhyloSift package \cite{darling_phylosift:_2014} to select gene markers for generating a phylogenetic tree and aligning them. Then, we concatenated the protein alignments for all 14 genomes and ran RaxML \cite{stamatakis_raxml_2014} with "PROTGAMMALG4M" amino acid substitution model, "a" algorithm and 100 alternative runs on distinct starting trees.

@@

For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database \cite{kanehisa_kegg:_2000}. This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2. Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class. Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
